Exploring Patterns in Clinical and Non-Clinical Listeria monocytogenes Isolates 

### 2025-04-21 code/analysis-code/statistical-analysis.qmd
updates to manuscript

---
title: 
author: 
date: 
output: html_document
---
# loading in packages
```{r}
library(tidyverse)
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
library(tidymodels)
library(lubridate)
library(ranger)

```
#Load in the data
```{r}
#Path to data. 
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

@@

```
```{r}

ggplot(mydata, aes(x = days_since_start, y = clinical_numeric)) +
  geom_jitter(height = 0.05, width = 5, alpha = 0.5) +  
  geom_line(data = mydata, aes(x = days_since_start, y = predicted_prob), color = "blue", size = 1) +
  labs(title = "Logistic Regression: Probability of Clinical Isolates Over Time",
       x = "Days Since Start",
       y = "Probability of Outcome (Yes = 1)") +
  theme_minimal()

```
```{r}
p <- ggplot(mydata, aes(x = days_since_start, fill = factor(clinical_numeric))) +  # Fill by outcome (0 or 1)
  geom_histogram(position = "identity", 
                 alpha = 0.5, 
                 bins = 30, 
                 color = "black") +  # Standard histogram
  labs(title = "Histogram of clinical isolates Over Time",
       x = "Days Since Start",
       y = "Count of Outcome (0 or 1)",
       fill = "Outcome") +
  theme_minimal()

figure_file=here("results/figures","clinicalxtime.png")
ggsave(filename=figure_file, plot = p, width = 8, height=6)
```

#More advanced temporal analysis of clinical isolates. Remember that every clinical isolate is like a human case of Listeriosis.
#For this analysis I will make some changes to my data by separating the time data into seasons.Winter is considered December, January and February; spring is March through May; summer is June through August; and fall or autumn is September through November.

@@

#Define the model
```{r}
rf_mod <-rand_forest(trees = 150, min_n =5) %>%
  set_mode("classification") %>%
  set_engine("ranger")
```
Create workflow
```{r}
workflow1 <- workflow() %>%
  add_model(rf_mod)%>%
  add_recipe(rf_recipe)
```
#fit the random forest model
```{r}
fit1 <- workflow1 %>%
  fit(data = train_data)
```
#Make predictions for evaluation
```{r}
predictions_rf <-fit1 %>%
  predict(new_data = test_data) %>%
  bind_cols(test_data)

@@

#Run metrics to evaluate the model
```{r}

# Create a metric set for classification
classification_metrics <- metric_set(accuracy, precision, recall, f_meas)

# Now use it on your predictions
metrics_rf <- predictions_rf %>%
  classification_metrics(truth = `Isolation type`, estimate = .pred_class)

@@

```{r}
metrics_rf <- predictions_rf %>%
  metrics(truth = `Isolation type`, estimate = .pred_class)
print(metrics_rf)
```
#Plot the results of this model
#Clean up some of the changes made to the data and add to processing file

>>>>

---
title: 
author: 
date: 
output: html_document
---
# loading in packages
```{r}
library(tidyverse)
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
library(tidymodels)
library(lubridate)
library(ranger)
library(parsnip)
library(poissonreg)
library(randomForest)
```
#Load in the data
```{r}
#Path to data. 
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

@@

# temporal analysis of clinical isolates. Remember that every clinical isolate is like a human case of Listeriosis.
#For this analysis I will make some changes to my data by separating the time data into seasons.Winter is considered December, January and February; spring is March through May; summer is June through August; and fall or autumn is September through November.

@@

#Define the model
```{r}
rf_mod <-rand_forest(trees = 150, min_n =5) %>%
  set_mode("classification") %>%
  set_engine("ranger",importance="impurity")
```
Create workflow
```{r}
workflow1 <- workflow() %>%
  add_model(rf_mod)%>%
  add_recipe(rf_recipe)
```
#fit the random forest model
```{r}
fit1 <- workflow1 %>%
  fit(data = train_data)
```
#Make predictions for evaluation
```{r}
predictions_rf <-fit1 %>%
  predict(new_data = test_data) %>%
  bind_cols(test_data)

@@

#Run metrics to evaluate the model
```{r}
# Create a metric set for classification
classification_metrics <- metric_set(accuracy, precision, recall, f_meas)
```

```{r}
#  use it on predictions
metrics_rf <- predictions_rf %>%
  classification_metrics(truth = `Isolation type`, estimate = .pred_class)

@@

```{r}
metrics_rf <- predictions_rf %>%
  metrics(truth = `Isolation type`, estimate = .pred_class)
print(metrics_rf)
```
#Plot the results of this model

```{r}
library(vip)  # for variable importance plot

# Extract the fitted model from the workflow
rf_fit <- extract_fit_parsnip(fit1)

# Plot variable importance
vip(rf_fit$fit)

#Poisson modeling using month

```{r}
# Create monthly clinical counts
monthly_counts <- mydata %>%
  filter(clinical == "Yes") %>%
  mutate(month = floor_date(`Create date`, "month")) %>%
  group_by(month) %>%
  summarize(clinical_count = n()) %>%
  ungroup()
#filling in gaps with 0
monthly_counts <- monthly_counts %>%
  complete(month = seq.Date(min(month), max(month), by = "month"),
           fill = list(clinical_count = 0))
#Feature engineering, extracting the month from the data
monthly_counts <- monthly_counts %>%
  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))

```{r}
library(tidymodels)
library(parsnip)
#Splitting the data 
set.seed(123)
data_split2 <- initial_time_split(monthly_counts, prop = 0.8)
train_data2 <- training(data_split2)
test_data2  <- testing(data_split2)

```
```{r}
#create a recipe 
poisson_recipe <- recipe(clinical_count ~ month_name, data = train_data2) %>%
  step_dummy(all_nominal_predictors())

```
```{r}
#Select the model
poisson_model <- 
  poisson_reg(
    mode = "regression",
    penalty = NULL,
    mixture = NULL,
    engine ="glm"
  )

```
```{r}
#Create a workflow 
library(poissonreg)
poisson_wf <-workflow() %>%
  add_model(poisson_model) %>%
  add_recipe(poisson_recipe)
```
```{r}
#fit the model
poisson_fit <- poisson_wf %>%
  fit(data = train_data2)
```
```{r}
#make predictions
predictions <- predict(poisson_fit, new_data = train_data2) %>%
  bind_cols(train_data2)
metrics(predictions, truth = clinical_count, estimate =.pred)
```
#Visualize results
```{r}
poisson_seasonal <- ggplot(predictions, aes(x = month)) +
  geom_line(aes(y = clinical_count), color = "black", linetype = "dashed") +
  geom_line(aes(y = .pred), color = "blue") +
  labs(title = "Predicted vs Actual Monthly Clinical Isolate Counts",
       y = "Clinical Count")
print(poisson_seasonal)
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/poisson_seasonal.png", plot = poisson_seasonal, width = 8, height = 6, dpi = 300)
```
#Predict future isolate counts based on seasonality
```{r}
last_month <-max(monthly_counts$month)
```

```{r}
future_months <- tibble(
  month = seq.Date(from = last_month %m+% months(1),by = "month", length.out = 12)
)%>%
  mutate(month_name = factor(month(month, label = TRUE), ordered= FALSE))
```

```{r}
#make predictions
future_predictions <- predict(poisson_fit, new_data = future_months)%>%
  bind_cols(future_months)
```
```{r}
#visualize
# Combine historical + future for plotting
combined_months<- bind_rows(
  monthly_counts %>% select(month, clinical_count) %>% mutate(source = "Actual"),
  future_predictions %>% rename(clinical_count = .pred) %>% mutate(source = "Forecast")
)

# Plot
seasonalpreds <- ggplot(combined_data, aes(x = month, y = clinical_count, color = source)) +
  geom_line() +
  labs(title = "Actual and Forecasted Clinical Isolate Counts",
       y = "Clinical Count",
       x = "Month") +
  scale_color_manual(values = c("Actual" = "black", "Forecast" = "blue")) +
  theme_minimal()
print(seasonalpreds)
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/seasonalpreds.png", plot = seasonalpreds, width = 8, height = 6, dpi = 300)
```

```{r}

# 1. Get the last month from your historical data
last_month <- max(monthly_counts$month)

# 2. Create the next 12 months
future_months <- tibble(
  month = seq.Date(from = last_month %m+% months(1), by = "month", length.out = 12)
) %>%
  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))

# 3. Make predictions using your Poisson model (or other model)
future_predictions <- predict(poisson_fit, new_data = future_months) %>%
  bind_cols(future_months)

# 4. Combine historical and forecast data
combined_months <- bind_rows(
  monthly_counts %>% 
    select(month, clinical_count) %>% 
    mutate(source = "Actual"),
  future_predictions %>% 
    rename(clinical_count = .pred) %>% 
    mutate(source = "Forecast")
)

# 5. Plot
seasonalpreds <- ggplot(combined_months, aes(x = month, y = clinical_count, color = source)) +
  geom_line(size = 1.2) +
  labs(
    title = "Actual and Forecasted Clinical Isolate Counts",
    y = "Clinical Count",
    x = "Month"
  ) +
  scale_color_manual(values = c("Actual" = "black", "Forecast" = "blue")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 6. Show plot
print(seasonalpreds)

# 7. Save to file
ggsave(
  "/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/seasonalpreds.png", 
  plot = seasonalpreds, width = 8, height = 6

```{r}
library(mgcv)
gam_model <- gam(clinical_count ~ s(day_of_year), family = poisson(), data = daily_counts)
plot(gam_model)
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/gam_model.png", plot = gam_model, width = 8, height = 6, dpi = 300)
```

```{r}
# Load necessary library
library(mgcv)
library(gratia) # for tidy ggplot GAM visualizations

# Example: Create ggplot-style plot from your GAM model
gam_plot <- draw(gam_model)  # `gam_model` is your fitted model

# Now save it
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/gam_model.png",
       plot = gam_plot, width = 8, height = 6, dpi = 300)

```{r}

#  Split the data into training and testing sets
set.seed(123)
data_split <- initial_split(monthly_counts, prop = 0.8)
train_data <- training(data_split)
test_data <- testing(data_split)

#  Define the recipe (clinical_count predicted by food_count)
poisson_recipe <- recipe(clinical_count ~ food_count, data = train_data)

# Define the Poisson regression model
poisson_model <- poisson_reg(mode = "regression", engine = "glm")

#  Create a workflow 
poisson_wf <- workflow() %>%
  add_recipe(poisson_recipe) %>%
  add_model(poisson_model)

#  Fit the model
poisson_fit <- poisson_wf %>%
  fit(data = train_data)

# Make predictions on the training data
predictions <- predict(poisson_fit, new_data = train_data) %>%
  bind_cols(train_data)

#  Evaluate the model 
isolate_metrics <- metrics(predictions, truth = clinical_count, estimate = .pred)
print(isolate_metrics)

# Visualize the results
ggplot(predictions, aes(x = month)) +
  geom_col(aes(y = clinical_count), fill = "lightgray", alpha = 0.8) +
  geom_line(aes(y = .pred), color = "blue", size = 1) +
  labs(
    title = "Poisson Regression: Predicting Clinical Isolates from Food Isolates",
    y = "Clinical Isolate Count",
    x = "Month"
  ) +
  theme_minimal()
```
#Chart showing likelihood
```{r}

#Now I want to do one more advanced statistical analysis. This analysis will utilize the counts of clinical and counts of food isolates each year and see if the amount of clinical isolates can be predicted by the amount of food isolates. This purpose of this analysis is to see if federal reguulations are doing a good job of preventing illness.

#put my food and clinical variable into counts and group by year

```{r}
monthly_counts2 <- mydata %>%
  mutate(month = floor_date(`Create date`, "month")) %>%
  count(month, isolation_source) %>%
  tidyr::pivot_wider(names_from = isolation_source, values_from = n, values_fill = 0)
```

```{r}
set.seed(123)  # for reproducibility

rf_model <- randomForest(
  clinical ~ food + environmental,
  data = monthly_counts2,
  ntree = 500,          # number of trees
  importance = TRUE     # track variable importance
)
print(rf_model)

```{r}
#model_predict_clinical <- lm(clinical ~ food + environmental, data = monthly_counts2)
#summary(model)

```{r}

`
#Visualzing the future
```{r}
# Suppose you created this:
future_data <- data.frame(
  month = as.Date(c("2025-02-01","2025-03-01","2025-04-01","2025-05-01", "2025-06-01", "2025-07-01","2025-08-01","2025-09-01","2025-10-01","2025-11-01","2025-12-01")),
  food = c(300,220, 180, 250,140,122,116,235,96,175,225),
  environmental = c(350,150, 170, 200,220,140,267,178,155,200,280)
)
```

```{r}
# Predict clinical isolates
future_data$Predicted_Clinical <- predict(rf_model, newdata = future_data)
```

```{r}
# Predict clinical isolates for future months
future_data$Predicted_Clinical <- predict(rf_model, newdata = future_data)

# Combine with past data (no predicted clinicals in the past, so we set it to NA)
combined_plot_data <- monthly_counts2 %>%
  select(month, clinical) %>%
  mutate(Predicted_Clinical = NA,
         Source = "Historical") %>%
  bind_rows(
    future_data %>%
      transmute(month, clinical = NA, Predicted_Clinical, Source = "Forecast")
  )

# Filter combined_plot_data for the last 3 years
recent_data <- combined_plot_data %>%
  filter(month >= (max(month) %m-% years(3)))

# Plot
ggplot(recent_data, aes(x = month)) +
  geom_line(aes(y = Predicted_Clinical, color = Source), size = 1.2, linetype = "dashed") +
  geom_line(data = recent_data %>% filter(!is.na(clinical)),
            aes(y = clinical, color = "Observed"), size = 1) +
  scale_color_manual(values = c("Observed" = "steelblue", "Historical" = "tomato", "Forecast" = "forestgreen")) +
  labs(title = "Observed and Forecasted Clinical Isolates (Past 3 Years)",
       x = "Month",
       y = "Count",
       color = "Legend") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r}
# Create new data (adjust as needed)
new_data <- data.frame(food = c(200), environmental = c(150))

# Predict clinical isolates
predict(model_predict_clinical, newdata = new_data)

```
#Machine Learning since this was just a simple linear regression
#Random forest
```{r}
#set seed and split data
set.seed(123)
data_split3 <- initial_split(model_data, prop = 0.8)
train_data3 <- training(data_split3)
test_data3 <- testing(data_split3)

### 2025-04-21 code/eda-code/eda.qmd
updates to manuscript

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

@@

#Making a table for the clinical variable 
```{r}
table(processeddata$clinical)
```
#doing the same for food
```{r}
table(processeddata$food)
```
#Now lets visualize these tables using bar graphs
```{r}
ggplot(mydata, aes(x = clinical)) + 
  geom_bar(fill = "steelblue") + 
  labs(title = "Yes/No Distribution", x = "Response", y = "Count") + 
  theme_minimal()

```
```{r}
library(ggplot2)

ggplot(mydata, aes(x = food)) + 
  geom_bar(fill = "orange") + 
  labs(title = "Yes/No Distribution", x = "Response", y = "Count") + 
  theme_minimal()

```
#cross tabulation
```{r}
table(mydata$food, mydata$clinical)
```
#We can also look at some distribution across the country since we have some state data
```{r}

mydata %>%
  filter(food == "Yes") %>%  # Keep only "Yes" cases
  count(Location) %>%  # Count occurrences per location
  ggplot(aes(x = Location, y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") + 
  labs(title = "Count of food isolated listeria Cases Across Locations", 
       x = "Location", 
       y = "Count of Yes") + 
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6,angle = 45, hjust = 1))  # Rotate x-axis labels for readability

```
#The left most bar is USA as a whole as some entries do not report the specific state they are from, the rest are states and territories. We see that NY has more cases than other states

@@

>>>>

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```
#Creating an additional variable called "Isolation_source" This is different from the `Isolation Source` variable in that it will include food and not just list it under environmental/other
```{r}
mydata <- mydata %>%
  mutate(
    clinical = as.character(clinical),
    food = as.character(food),
    isolation_source = case_when(
      clinical == "Yes" ~ "clinical",
      food == "Yes" ~ "food",
      TRUE ~ "environmental"
    )
  )

```

@@

#Making a table for the clinical variable 
```{r}
table(mydata$clinical)
```
#doing the same for food
```{r}
table(mydata$food)
```
```{r}
table(mydata$isolation_source)
```

#Now lets visualize these tables using bar graphs of counts for food clincial and env
```{r}

isolatetypes <- ggplot(mydata, aes(x = isolation_source)) +
  geom_bar(fill = "seagreen") +
  labs(title = "Isolation Source Distribution",
       x = "Isolation Source",
       y = "Count") +
  theme_minimal()
 print(isolatetypes)  
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/isolatetypes.png", plot = isolatetypes, width = 8, height = 6, dpi = 300)
```
#As we see here, clinical cases are rare but that does not take away from the fact that they are usually severe.
```{r}
#cross tabulation
table(mydata$food, mydata$clinical)
```
#We can also look at some distribution across the country since we have some state data

```{r}

locations <- ggplot(
  mydata %>% 
    filter(food == "Yes", Location != "USA"),
  aes(x = Location)
) + 
  geom_bar(fill = "steelblue") + 
  labs(
    title = "Count of Food-Isolated Listeria Cases Across Locations",
    x = "Location",
    y = "Count of Yes"
  ) + 
  theme_minimal() + 
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1))
print(locations)
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/locations.png", plot = locations, width = 8, height = 6, dpi = 300)
```

```{r}
locations2 <- ggplot(mydata %>% filter(clinical == "Yes"), aes(x = Location)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Count of Clincally-Isolated Listeria Cases Across Locations",
       x = "Location",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1))
print(locations2)
```
#This only tells me that all cases of listeriosis were in the U.S. which i specified for so we wont use this graph.

#The left most bar is USA as a whole as some entries do not report the specific state they are from, the rest are states and territories. We see that NY has more cases than other states
```{r}
sourcestate <- mydata %>%
  filter(clinical == "Yes", !is.na(`Food origin`)) %>%  
  count(`Food origin`) %>%  
  ggplot(aes(x = `Food origin`, y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Count of Clinically Isolated Listeria Sources by Food Origin",
       x = "Food Origin",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1))

print(sourcestate)
```

```{r}
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/sourcestate.png", plot = sourcestate, width = 8, height = 6, dpi = 300)
```
#Based on this graph we see that many of the clinical cases are coming from foods produced in Sweden...interesting. maybe we can use food origin as a predictor as well.

#The left most bar is USA as a whole as some entries do not report the specific state they are from, the rest are states and territories. We see that NY has more cases than other states

@@

#plot of food isolates and clinical isolates across 10 years
```{r}
library(tidyverse)
library(lubridate)

# Step 1: Prepare the data — monthly counts
monthly_counts <- mydata %>%
  mutate(month = floor_date(`Create date`, "month")) %>%
  group_by(month) %>%
  summarize(
    food_count = sum(food == "Yes"),
    clinical_count = sum(clinical == "Yes")
  ) %>%
  ungroup()

# Step 2: Convert to long format for plotting
monthly_long <- monthly_counts %>%
  pivot_longer(cols = c(food_count, clinical_count),
               names_to = "source_type",
               values_to = "count")

# Step 3: Plot
ggplot(monthly_long, aes(x = month, y = count, color = source_type)) +
  geom_line(size = 1.2) +
  labs(
    title = "Monthly Salmonella Isolate Counts by Source",
    x = "Month",
    y = "Isolate Count",
    color = "Source Type"
  ) +
  scale_color_manual(values = c("food_count" = "forestgreen", "clinical_count" = "firebrick")) +
  theme_minimal()
```

```{r}
Isolatecounts <- ggplot(monthly_long, aes(x = month, y = count, fill = source_type)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Monthly Salmonella Isolate Counts by Source (Stacked Bars)",
    x = "Month",
    y = "Isolate Count",
    fill = "Source Type"
  ) +
  scale_fill_manual(values = c("food_count" = "forestgreen", "clinical_count" = "firebrick")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/isolatecounts.png", plot = Isolatecounts, width = 8, height = 6, dpi = 300)
print(Isolatecounts)
```

### 2025-04-21 products/manuscript/Manuscript.qmd
updates to manuscript

---
title: "Exploring clinical and non-clinical(food) Listeria monocytogenes isolates "
author: "Alexis Gonzalez"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

@@

Will complete at the end

@@

Listeria monocytogenes is a ubiquitous, gram-positive, intracellular pathogen renowned for its adaptability and resilience\[5\]. This bacterium can be found in a wide range of environments such as soil, water, animals, and vegetation. Due to its ability to proliferate in food products and low temperatures, it is considered a major food-borne illness. The ingestion of Listeria-infected food may lead to listeriosis, a gastrointestinal illness that causes vomiting, diarrhea, and even death. It is of even greater concern in pregnant individuals, as its forceful invasion tactics allow the bacterium to cross the placental barrier, infecting the fetus; this can lead to miscarriages and stillbirths\[10\]. While immunocompetent individuals may only experience mild gastroenteritis, disseminated infection in the immunocompromised has a lethality rate of \~30%\[8\].  Listeriosis causes 1,600 infections and 260 deaths each year with a 95% hospitalization rate. The United States Department of Agriculture lists 6 foods that are at high risks for L.monocytogenes contamination; unpasteurized dairy products, soft cheeses, raw fruits and vegetables (particularly sprouts, enoki mushrooms, and melons), Ready-To-Eat deli meats and hot dogs, refrigerated pate, and refrigerated smoked seafood. Listeria contamination typically happens post-lethality which exposes a ready-to-eat product to a processing environment after a form of lethality treatment has taken place. Post lethality processing environments can be deli slicers, submersion in a brine solution, meat encassing, bagging or packaging. The USDA recommends all establishments to take part in additional post lethality treatment such as antimicrobial coatings, pasteurization, or chemical or physical sanitation of equipment. Sanitation tends to be the more cost effective option for many establishments but, due to incorrect sanitation practices and L.monocytogenes robust bio-film formation capabilities sanitation does not always prevent contamination. Outbreaks have been seen in a range of products in recent years. From 2023 to 2025 there have been over 10 multi-state listeria outbreaks. The most recent being an outbreak beginning February 22, 2025 in frozen supplement shakes. These shakes were distributed among institutional facilities such as long term care facilities and hospitals. This outbreak has resulted in 38 cases, 37 hospitalizations, and 12 deaths so far as this investigation is ongoing as of April 2025.

\
This study aims to investigate trends in listeria isolates found in clinical and non-clinical samples in efforts to predict cases.  This study will investigate time of a year and number of non-clinical isolates as a predictor for clinical isolates, and food type as a predictor for food isolates , non-clinical isolates that originate in food products. The data source being used to conduct this study is the National Center for Biotechnology Information's Pathogen Detection browser. Pathogen Detection is a culmination of pathogen genomic sequences from federal, state, private, and foreign laboratories. It is utilized constantly by public health officials,epidemiologists, and scientists to identify chains of transmission. The new isolates are analyzed in real time to aid in that identification by developing clusters. Listeria monocytogenes is a nationally notifiable disease so, many laboratories opt to share their data nationally through the Pathogen Detection Browser. By utilizing the most up to date data and identifying trends, this study hopes to provide information to form new guidelines on protecting the public from Listeriosis infection.

@@

@@

## Exploratory/Descriptive analysis

```{r}
knitr::include_graphics(here("results/figures", "foodisolatecount.png"))
```

## Basic statistical analysis

## Full analysis

#Need to insert figures

```{r}
knitr::include_graphics(here("results/figures", "clinicalxtime.png"))

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."...................;;;;;;;;;;;;;;;::::::::::::::::P
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

@@

@@

@@

@@

>>>>

---
title: "Exploring Patterns in Clinical and Non-Clinical Listeria monocytogenes Isolates "
author: "Alexis Gonzalez"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

@@

Listeria monocytogenes is major food born pathogen. Though infections are rare, L.monocytogenes has the highest hospitalization rate and death rate of the common food borne illnesses. This pathogen can proliferate in a number of hosts and climates. Understanding temporal and source relationships may allow the public to be better informed on how to protect themselves. This study used whole-genome sequencing data from the NCBI Pathogen Detection database (2010--2024) to explore trends in food, clinical, and environmental isolates of *L. monocytogenes*. We conducted exploratory data analysis to assess the frequency and geographic distribution of isolates. Poisson and logistic regression models were applied to evaluate the effect of seasonality on clinical isolate occurrence. 33,264 isolates were analyzed, with a majority of them being environmental. Geographic analysis concluded that New York state contributed a large percentage of food isolates, and many illness-causing foods originated in Sweden. Statistical modeling showed a consistent seasonal trends. Random forest models concluded that days of the year were an important factor and Poisson regression allowed for prediction of future clinical isolates based on the number of environmental and food isolates, though the association was weak. These findings suggest that utilizing temporal relationships may be effective in predicting potential outbreaks.

@@

Listeria monocytogenes is a ubiquitous, gram-positive, intracellular pathogen renowned for its adaptability and resilience\[5\]. This bacterium can be found in a wide range of environments such as soil, water, animals, and vegetation. Due to its ability to proliferate in food products and low temperatures, it is considered a major food-borne illness. The ingestion of Listeria-infected food may lead to listeriosis, a gastrointestinal illness that causes vomiting, diarrhea, and even death. It is of even greater concern in pregnant individuals, as its forceful invasion tactics allow the bacterium to cross the placental barrier, infecting the fetus; this can lead to miscarriages and stillbirths\[10\]. While immunocompetent individuals may only experience mild gastroenteritis, disseminated infection in the immunocompromised has a lethality rate of \~30%\[8\].  Listeriosis causes 1,600 infections and 260 deaths each year with a 95% hospitalization rate. The United States Department of Agriculture lists 6 foods that are at high risks for L.monocytogenes contamination; unpasteurized dairy products, soft cheeses, raw fruits and vegetables (particularly sprouts, enoki mushrooms, and melons), Ready-To-Eat deli meats and hot dogs, refrigerated pate, and refrigerated smoked seafood. Listeria contamination typically happens post-lethality which exposes a ready-to-eat product to a processing environment after a form of lethality treatment has taken place. Post lethality processing environments can be deli slicers, submersion in a brine solution, meat encasing, bagging or packaging. The USDA recommends all establishments to take part in additional post lethality treatment such as antimicrobial coatings, pasteurization, or chemical or physical sanitation of equipment. Sanitation tends to be the more cost effective option for many establishments but, due to incorrect sanitation practices and L.monocytogenes robust bio-film formation capabilities sanitation does not always prevent contamination. Outbreaks have been seen in a range of products in recent years. From 2023 to 2025 there have been over 10 multi-state listeria outbreaks. The most recent being an outbreak beginning February 22, 2025 in frozen supplement shakes. These shakes were distributed among institutional facilities such as long term care facilities and hospitals. This outbreak has resulted in 38 cases, 37 hospitalizations, and 12 deaths so far as this investigation is ongoing as of April 2025.

\
This study aims to investigate trends in listeria isolates found in clinical and non-clinical samples in efforts to predict cases.  This study will investigate time of year and number of non-clinical isolates as a predictor for clinical isolates, and food type as a predictor for food isolates , non-clinical isolates that originate in food products. The data source being used to conduct this study is the National Center for Biotechnology Information's Pathogen Detection browser. Pathogen Detection is a culmination of pathogen genomic sequences from federal, state, private, and foreign laboratories. It is utilized constantly by public health officials,epidemiologists, and scientists to identify chains of transmission. The new isolates are analyzed in real time to aid in that identification by developing clusters. Listeria monocytogenes is a nationally notifiable disease so, many laboratories opt to share their data nationally through the Pathogen Detection Browser. By utilizing the most up to date data and identifying trends, this study hopes to provide information to form new guidelines on protecting the public from Listeriosis infection.

@@

Two additional columns were created for exploratory analysis to separate clinical and non clinical isolates from the "Isolation type variable". Entries for the "Clinical" column were coded as "Yes" if the entry for "Isolation type" was clinical. All non-clinical isolates were coded as "environmental/other" so the "Isolation Source" column was utilized to create the "Food" column. All isolates that came from various meat and dairy, and produce food sources were coded as "Yes". Additional columns created were "days_since_start" which converted all dates in the selected time period into days since the first date in the data set. A column for year, day, month, and season were also created for statistical analysis.

For temporal analysis was conducted using poisson regression. All machine learning models were conducted using the tidymodels workflow. Specific packages for poisson modeling were parsnip and poissonreg.

@@

### Exploratory Data Analysis

Exploratory data analysis started with visualizing the total counts of food, clinical and environmental isolates in the 10 year time period. It is seen in *figure* 1 that food isolates account for the smallest proportion of isolated *Listeria monocytogenes* at 10,284. Clinical isolates fall slightly ahead at 10,319, with a difference of 35 isolates in the 10 year period. Environmental samples dominate the isolate types at 12,777. To get a visualization of where food isolates were coming from a bar graph was made showing the location of origin of each food isolate. Some isolates only logged "USA" as their origin so those were removed from visualization. The majority of food sampled were originate in New York state, followed by Washington, Arizona, Florida, and Texas. The 3rd bar graph *figure 3* shows the origin of clinical cases. This means the origin of the source of infection, likely a food product. An overwhelming percentage of originated in Sweden, followed by Texas and New York.

```{r}
knitr::include_graphics(here("results/figures", "isolatetypes.png"))
```

```{r}
knitr::include_graphics(here("results/figures","locations.png"))
```

```{r}
knitr:: include_graphics(here("results/figures", "sourcestate.png"))
```

```{r}
knitr::include_graphics(here("results/figures", "isolatecounts.png"))
```

## Temporal Analysis

*Time of year or "season" as a predictor for clinical Listeria monocytogenes isolates*

The first goal of this analysis is to find any seasonal patterns in *Listeria* infections and *Listeria* contaminated food items.

```{r}
knitr::include_graphics(here("results/figures", "gam_model.png"))
```

```{r}
knitr::include_graphics(here("results/figures", "seasonalpreds.png"))
```

```{r}
knitr::include_graphics(here("results/figures", "poisson_seasonal.png"))
```

@@

## Logistic Regression

@@

This study examined temporal and source-based trends in Listeria monocytogenes isolates reported in the United States from 2010 to 2024 using publicly available data from the NCBI Pathogen Detection database. Isolates were classified into clinical, environmental, and food-related categories to assess patterns and predictors of clinical infections.

Exploratory analysis showed that environmental samples were the most frequently reported isolate type, which supports previous findings that L. monocytogenes is widespread in processing environments and natural reservoirs. Clinical and food isolates were nearly equal in count, suggesting a potential link between food exposure and clinical outcomes.

Temporal modeling using Poisson regression revealed a seasonal pattern in clinical isolates, with elevated counts in warmer months---especially summer. This pattern may reflect greater bacterial growth in food products stored at warmer temperatures, seasonal shifts in food handling practices, or increased consumption of ready-to-eat items during these months. Logistic regression analysis reinforced these findings, highlighting seasonality as a potential predictor of clinical listeriosis.

These results align with prior epidemiological data suggesting that Listeria infections may follow seasonal trends, which has important implications for food safety monitoring and public health messaging. Summer could be a critical time for intensified surveillance and intervention efforts.

@@

While the data set contains thousands of isolates, the amount of clinical isolates is quite small due to the rarity of infection. This made statistical analysis complicated as some months during the 10 year period experienced 0 cases of listeriosis. There also were not many predictors to work with from the raw data. Isolation source would have been an association to explore but, the isolation sources are entered manually bu researchers and inspectors so there is no standardization in nomenclature of sources. For example, one source can say "hot-dog" while another says "premium hot-dog". More spatial analysis could have been done if more entries specify the state of origin. Many had just "USA" or NA. This is also not standardized for NCBI and can be entered however the submitter pleases.

@@

This study identified seasonal patterns in clinical Listeria monocytogenes isolates, with potential implications for timing surveillance and prevention strategies. While environmental and food isolates remain a significant reservoir for transmission, more refined metadata and consistent data entry practices are essential for improving future predictive models.

Future research should focus on expanding spatial analysis, refining source categorization through natural language processing, and incorporating additional predictors such as food recall data or consumer behavior trends. With enhanced data quality and analytical tools, real-time surveillance systems can be better leveraged to prevent and control outbreaks of listeriosis.
